A Speculative Approach to Parallelization in Particle Swarm Optimization

### 2011-09-09 masters_thesis/si_paper/paper.tex
suggestions for making plots more readable (I only changed sphere20.gpi)
Done with reviewer comments, and so probably completely done with the paper

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{iters_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of iterations of the algorithm performed.}
  \label{fig:iters-sphere}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes, showing performance after
  a set number of function evaluations.  Error bars show mean and 10th and 90th
  percentiles.}
  \label{fig:evals-sphere}
\end{figure}

@@

\section{Relaxing the Requirements}
\label{sec:relax}

@@

We outline three main improvements to speculative evaluation.  First, in
\sec{pickbest} we describe a method that uses all of the information found in
doing speculative evaluations.  Then Section~\ref{sec:pruning} presents a
technique that reduces the number of speculative evaluations that need to be
done for each particle, allowing speculative evaluation to use larger swarm
sizes with the same number of processors.  Finally, \sec{manyiters} shows a
method for speculating several iterations ahead, instead of just one.

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{sphere20}
  \caption{Function Sphere with 20 dimensions, comparing a pertinent subset of
  possible methods.  Each method performs one evaluation on each of 240
  processors per time step.}
  \label{fig:sphere-20}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{rastrigin}
  \caption{Function Rastrigin with 20 dimensions.  Each method performs one
  evaluation on each of 240 processors per time step.}
  \label{fig:rastrigin}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank6}
  \caption{Function Griewank with 20 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.  Instead of showing
  average function value, we show the percentage of runs that have found the
  global optimum by each iteration.  All algorithms use the Ring topology.}
  \label{fig:basic-griewank1}
\end{figure}

@@

We show results in \fig{basic-griewank2} for swarms of size 30 and 240 using
the Ring topology.  One can see in the figure that when SEPSO is successful, it
finds the optimum much faster than Standard.  However, because the swarm size
is so small, SEPSO gets stuck a little less than half of the time.

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank1}
  \caption{Function Griewank with 20 dimensions.  Each method performs one
  evaluation of the objective function on each of 240 processors per time step.
  All algorithms use the Ring topology.}
  \label{fig:basic-griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank3}
  \caption{Function Griewank with 20 dimensions, comparing Pick Best with
  results from \fig{basic-griewank2}.  Each method performs one evaluation on
  each of 240 processors per time step.  All algorithms use the Ring topology.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank4}
  \caption{Function Griewank with 20 dimensions, comparing pruning with the
  best results from \fig{griewank-pickbest}.  Each method performs one
  evaluation on each of 240 processors per time step.  All algorithms use the
  Ring topology.}
  \label{fig:griewank-pruned}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank5}
  \caption{Function Griewank with 20 dimensions, comparing speculating several
  iterations ahead with selections from \fig{griewank-pruned}.  Each method
  performs one evaluation on each of 240 processors per time step.  All
  algorithms use the Ring topology.}
  \label{fig:griewank-manyiters}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewankall}
  \caption{Function Griewank with 20 dimensions, comparing the performance of
  standard and speculative parallelizations as the number of processors
  increases.  All algorithms use the Ring topology.}
  \label{fig:griewank-all}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{bohachevsky}
  \caption{Function Bohachevsky with 20 dimensions.  Each method performs one
  evaluation on each of 480 processors per time step.}
  \label{fig:bohachevsky}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{sphere500}
  \caption{Function Sphere with 500 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.}
  \label{fig:sphere500}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{griewank500}
  \caption{Function Griewank with 500 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.}
  \label{fig:griewank500}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{rastrigin500}
  \caption{Function Rastrigin with 500 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.}
  \label{fig:rastrigin500}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{rbf}
  \caption{Results for fitting a radial basis function to noisy data.  We use
  144 processors for each method, so each time step corresponds to 144 function
  evaluations.}
  \label{fig:rbf}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{rbf_stuck}
  \caption{Results for fitting a radial basis function to noisy data.  We use
  144 processors for each method, so each time step corresponds to 144 function
  evaluations.}
  \label{fig:rbf-stuck}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.7\columnwidth]{bare_bones}
  \caption{Results for optimizing the 20-dimensional Sphere function with Bare
  Bones PSO.  We use 240 processors for each method, so each time step
  corresponds to 240 function evaluations.}
  \label{fig:bare-bones}
\end{figure}

@@

What we have presented is not a new variant of PSO.  We presented a new
parallelization technique, so we compared parallelization strategies on the
same algorithm, the original PSO.  Our most promising techniques do change the
behavior of the PSO algorithm slightly, as do some other previously proposed
parallelization techniques, such as asynchronous PSO.  However, our methods are
applicable to almost all PSO variants, and so a comparison using the same
variant for each of the parallelization techniques tested is justified.  We
presented results on a separate PSO variant that give evidence for this claim.

>>>>

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{iters_sphere}
  \caption{Function Sphere with various swarm sizes, comparing performance with
  the number of iterations of the algorithm performed.}
  \label{fig:iters-sphere}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{evals_sphere}
  \caption{Function Sphere with various swarm sizes, showing performance after
  a set number of function evaluations.  Error bars show mean and 10th and 90th
  percentiles.}
  \label{fig:evals-sphere}
\end{figure}

@@

\section{Speculative Variations on PSO}
\label{sec:relax}

@@

We outline three main improvements to speculative evaluation.  First, in
\sec{pickbest} we describe a method that uses all of the information found in
doing speculative evaluations.  Then \sec{pruning} presents a technique that
reduces the number of speculative evaluations that need to be done for each
particle, allowing speculative evaluation to use larger swarm sizes with the
same number of processors.  Finally, \sec{manyiters} shows a method for
speculating several iterations ahead, instead of just one.

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{sphere20}
  \caption{Function Sphere with 20 dimensions, comparing a pertinent subset of
  possible methods.  Each method performs one evaluation on each of 240
  processors per time step.}
  \label{fig:sphere-20}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{rastrigin}
  \caption{Function Rastrigin with 20 dimensions.  Each method performs one
  evaluation on each of 240 processors per time step.}
  \label{fig:rastrigin}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{griewank6}
  \caption{Function Griewank with 20 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.  Instead of showing
  average function value, we show the percentage of runs that have found the
  global optimum by each iteration.  All algorithms use the Ring topology.}
  \label{fig:basic-griewank1}
\end{figure}

@@

We show results in \fig{basic-griewank2} for swarms of size 30 and 240 using
the Ring topology.  One can see in the figure that when SEPSO is successful, it
finds the optimum much faster than Standard.  However, because the swarm size
is so small, SEPSO gets stuck almost half of the time.

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{griewank1}
  \caption{Function Griewank with 20 dimensions.  Each method performs one
  evaluation of the objective function on each of 240 processors per time step.
  All algorithms use the Ring topology.}
  \label{fig:basic-griewank2}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{griewank3}
  \caption{Function Griewank with 20 dimensions, comparing Pick Best with
  results from \fig{basic-griewank2}.  Each method performs one evaluation on
  each of 240 processors per time step.  All algorithms use the Ring topology.}
  \label{fig:griewank-pickbest}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{griewank4}
  \caption{Function Griewank with 20 dimensions, comparing pruning with the
  best results from \fig{griewank-pickbest}.  Each method performs one
  evaluation on each of 240 processors per time step.  All algorithms use the
  Ring topology.}
  \label{fig:griewank-pruned}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{griewank5}
  \caption{Function Griewank with 20 dimensions, comparing speculating several
  iterations ahead with selections from \fig{griewank-pruned}.  Each method
  performs one evaluation on each of 240 processors per time step.  All
  algorithms use the Ring topology.}
  \label{fig:griewank-manyiters}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{griewankall}
  \caption{Function Griewank with 20 dimensions, comparing the performance of
  standard and speculative parallelizations as the number of processors
  increases.  All algorithms use the Ring topology.}
  \label{fig:griewank-all}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{bohachevsky}
  \caption{Function Bohachevsky with 20 dimensions.  Each method performs one
  evaluation on each of 480 processors per time step.}
  \label{fig:bohachevsky}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{sphere500}
  \caption{Function Sphere with 500 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.}
  \label{fig:sphere500}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{griewank500}
  \caption{Function Griewank with 500 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.}
  \label{fig:griewank500}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{rastrigin500}
  \caption{Function Rastrigin with 500 dimensions.  Each method performs one
  evaluation on each of 800 processors per time step.}
  \label{fig:rastrigin500}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{rbf}
  \caption{Results for fitting a radial basis function to noisy data.  We use
  144 processors for each method, so each time step corresponds to 144 function
  evaluations.}
  \label{fig:rbf}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{rbf_stuck}
  \caption{Results for fitting a radial basis function to noisy data.  We use
  144 processors for each method, so each time step corresponds to 144 function
  evaluations.}
  \label{fig:rbf-stuck}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[width=.9\columnwidth]{bare_bones}
  \caption{Results for optimizing the 20-dimensional Sphere function with Bare
  Bones PSO.  We use 240 processors for each method, so each time step
  corresponds to 240 function evaluations.}
  \label{fig:bare-bones}
\end{figure}

@@

What we have presented is not a new variant of the PSO equations.  We presented
a new parallelization technique, so we compared parallelization strategies on
the same algorithm, the original PSO.  Our most promising techniques do change
the behavior of the PSO algorithm slightly, as do some other previously
proposed parallelization techniques, such as asynchronous PSO.  However, our
methods are applicable to almost all PSO variants, and so a comparison using
the same variant for each of the parallelization techniques tested is
justified.  We presented results on a separate PSO variant that give evidence
for this claim.
